---
title: "Project 5"
author: "Aleksandra Buchowicz, Michał Klimek, Marcin Ostrowski"
date: "2024-01-11"
output:
  beamer_presentation: default
  slidy_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)

library(ggplot2)
library(patchwork)
library(tidyr)
library(glmnet)
library(pROC)
```

# Loading the data

```{r data_head}
data <- read.csv('data_project_5_for_students.csv')
head(data)
```

# Basic statistics
First, let's get a basic data summary.

```{r data_summary}
summary(data)
```
We can create tables to see, what values are in each column.

```{r tables, echo=FALSE}
for (col in colnames(data)){
  print(table(data[,col], useNA="always"))
}
```

We can see that there is some missing data in the dataset. The total number of missing data instances in the entire data set:

```{r nas}
sum(is.na(data))
```
We have also created plots to summarize distributions of data in each column.

```{r plot_all}
p00 <- ggplot(data, aes(x=Group)) +
  geom_bar(stat="count", width=0.7)+
  theme_minimal()
p0 <- ggplot(data, aes(x=Gender)) +
  geom_bar(stat="count", width=0.7)+
  theme_minimal()
p1 <- ggplot(data, aes(x=rs2476601)) +
  geom_bar(stat="count", width=0.7)+
  theme_minimal()
p2 <- ggplot(data, aes(x=rs3087243)) +
  geom_bar(stat="count", width=0.7)+
  theme_minimal()
p3 <- ggplot(data, aes(x=rs3807306)) +
  geom_bar(stat="count", width=0.7)+
  theme_minimal()
p4 <- ggplot(data, aes(x=rs1800629)) +
  geom_bar(stat="count", width=0.7)+
  theme_minimal()
p5 <- ggplot(data, aes(x=rs1799724)) +
  geom_bar(stat="count", width=0.7)+
  theme_minimal()

p00+ p0 + p1 + p2 + p3 + p4 + p5 + plot_annotation(title = "Distributions of each column in the dataset")
```
# Distributions of data among CFS and control groups

```{r group_hist, echo = FALSE}

p1 <- ggplot(data, aes(x=rs2476601, fill=Group)) +
  geom_bar(stat="count", width=0.7)+
  theme_minimal()
p2 <- ggplot(data, aes(x=rs3087243, fill=Group)) +
  geom_bar(stat="count", width=0.7)+
  theme_minimal()
p3 <- ggplot(data, aes(x=rs3807306, fill=Group)) +
  geom_bar(stat="count", width=0.7)+
  theme_minimal()
p4 <- ggplot(data, aes(x=rs1800629, fill=Group)) +
  geom_bar(stat="count", width=0.7)+
  theme_minimal()
p5 <- ggplot(data, aes(x=rs1799724, fill=Group)) +
  geom_bar(stat="count", width=0.7)+
  theme_minimal()

p1 + p2 + p3 + p4 + p5 + plot_annotation(title = "Distribution of genes among both groups")

```

## CFS group

```{r cfs_tables, echo=FALSE}
print("### Chronic Fatigue Syndrome ###")
for (col in colnames(data)[colnames(data)!="Group"]){
  print(table(data[data$Group=='CFS',col], useNA="always"))
}
```
```{r cfs_hist, echo = FALSE}
library(patchwork)
data_cfs <- data[data$Group=='CFS',]

p1 <- ggplot(data_cfs, aes(x=rs2476601, fill=Gender)) +
  geom_bar(stat="count", width=0.7)+
  theme_minimal()
p2 <- ggplot(data_cfs, aes(x=rs3087243, fill=Gender)) +
  geom_bar(stat="count", width=0.7)+
  theme_minimal()
p3 <- ggplot(data_cfs, aes(x=rs3807306, fill=Gender)) +
  geom_bar(stat="count", width=0.7)+
  theme_minimal()
p4 <- ggplot(data_cfs, aes(x=rs1800629, fill=Gender)) +
  geom_bar(stat="count", width=0.7)+
  theme_minimal()
p5 <- ggplot(data_cfs, aes(x=rs1799724, fill=Gender)) +
  geom_bar(stat="count", width=0.7)+
  theme_minimal()

p1 + p2 + p3 + p4 + p5 + plot_annotation(title = "CFS Group")

```

## Healthy control group

```{r hc_tables, echo=FALSE}
print("### Healthy Control ###")
for (col in colnames(data)[colnames(data)!="Group"]){
  print(table(data[data$Group=='HC',col], useNA="always"))
}
```

```{r hc_hist, echo = FALSE}
library(patchwork)
data_hc <- data[data$Group=='HC',]

p1 <- ggplot(data_hc, aes(x=rs2476601, fill=Gender)) +
  geom_bar(stat="count", width=0.7)+
  theme_minimal()
p2 <- ggplot(data_hc, aes(x=rs3087243, fill=Gender)) +
  geom_bar(stat="count", width=0.7)+
  theme_minimal()
p3 <- ggplot(data_hc, aes(x=rs3807306, fill=Gender)) +
  geom_bar(stat="count", width=0.7)+
  theme_minimal()
p4 <- ggplot(data_hc, aes(x=rs1800629, fill=Gender)) +
  geom_bar(stat="count", width=0.7)+
  theme_minimal()
p5 <- ggplot(data_hc, aes(x=rs1799724, fill=Gender)) +
  geom_bar(stat="count", width=0.7)+
  theme_minimal()

p1 + p2 + p3 + p4 + p5 + plot_annotation("Healthy Control")

```

# Distributions of data between genders

## Distribution of genes in females

```{r female_tables, echo=FALSE}
print("### Females ###")
for (col in colnames(data)[colnames(data)!="Gender"]){
  print(table(data[data$Gender=='f',col], useNA="always"))
}
```

```{r f_hist, echo = FALSE}
library(patchwork)
data_f <- data[data$Gender=='f',]

p1 <- ggplot(data_f, aes(x=rs2476601, fill=Group)) +
  geom_bar(stat="count", width=0.7)+
  theme_minimal()
p2 <- ggplot(data_f, aes(x=rs3087243, fill=Group)) +
  geom_bar(stat="count", width=0.7)+
  theme_minimal()
p3 <- ggplot(data_f, aes(x=rs3807306, fill=Group)) +
  geom_bar(stat="count", width=0.7)+
  theme_minimal()
p4 <- ggplot(data_f, aes(x=rs1800629, fill=Group)) +
  geom_bar(stat="count", width=0.7)+
  theme_minimal()
p5 <- ggplot(data_f, aes(x=rs1799724, fill=Group)) +
  geom_bar(stat="count", width=0.7)+
  theme_minimal()

p1 + p2 + p3 + p4 + p5 + plot_annotation(title = "Distribution of genes in females")

```

## Distribution of genes in males

```{r male_tables}
print("### Males ###")
for (col in colnames(data)[colnames(data)!="Gender"]){
  print(table(data[data$Gender=='m',col], useNA="always"))
}
```


```{r m_hist, echo = FALSE}
library(patchwork)
data_m <- data[data$Gender=='m',]

p1 <- ggplot(data_m, aes(x=rs2476601, fill=Group)) +
  geom_bar(stat="count", width=0.7)+
  theme_minimal()
p2 <- ggplot(data_m, aes(x=rs3087243, fill=Group)) +
  geom_bar(stat="count", width=0.7)+
  theme_minimal()
p3 <- ggplot(data_m, aes(x=rs3807306, fill=Group)) +
  geom_bar(stat="count", width=0.7)+
  theme_minimal()
p4 <- ggplot(data_m, aes(x=rs1800629, fill=Group)) +
  geom_bar(stat="count", width=0.7)+
  theme_minimal()
p5 <- ggplot(data_m, aes(x=rs1799724, fill=Group)) +
  geom_bar(stat="count", width=0.7)+
  theme_minimal()

p1 + p2 + p3 + p4 + p5 + plot_annotation(title = "Distribution of genes in males")

```

# Distributions of genetic markers in healthy patients

Information about the the respective genetic information and the allele distribution of different populations comes from the 1000 Human Genome Project. We have retrieved the information with the use of ensembl.org.

The information about the genetic markers will be presented using the VCF format: `(CHROM POS ID REF ALT)`, where:\newline

* `CHROM` - chromosome identifier
* `POS` - reference position
* `ID` - genetic marker identifier
* `REF` - reference base(s) 
* `ALT` - alternative base(s)


# rs2476601

The information about this genetic marker can be found here: 
https://www.ensembl.org/Homo_sapiens/Variation/Population?db=core;r=1:113834446-113835446;v=rs2476601;vdb=variation;vf=166892

VCF: `1  113834946  rs2476601  A  G,T` 

```{r rs2476601_data, echo=FALSE}
region <- c("ALL", "AFR", "AMR", "EAS", "EUR", "SAS")
A_freq <- c(0.027, 0.003, 0.036, 0.000, 0.094, 0.013)
G_freq <- c(0.973, 0.997, 0.964,  1.000, 0.906, 0.987)

AA_freq <- c(0.001 , 0.00, 0.003 ,  0.000, 0.004 , 0.00)
AG_freq <- c(0.052 , 0.006 , 0.066 ,  0.000, 0.181 , 0.027)
GG_freq <- c(0.946 , 0.994 , 0.931 ,  1.000, 0.815 , 0.973)

data.rs2476601.rep <- data.frame(region, A_freq, G_freq, AA_freq, AG_freq, GG_freq)
print(data.rs2476601.rep)
```
# rs3087243

The information about this genetic marker can be found here: 
https://www.ensembl.org/Homo_sapiens/Variation/Population?db=core;r=2:203873696-203874696;v=rs3087243;vdb=variation;vf=89583355

VCF: `2  203874196  rs3087243  G  A,T` 

```{r rs3087243_data, echo=FALSE}
region <- c("ALL", "AFR", "AMR", "EAS", "EUR", "SAS")
A_freq <- c(0.369 , 0.176 , 0.380 , 0.263 , 0.470 , 0.627)
G_freq <- 1.00 - A_freq

GG_freq <- c(0.421, 0.670, 0.372,  0.548, 0.274, 0.139)
AA_freq <- c(0.159, 0.023, 0.133,  0.073, 0.215, 0.393)
AG_freq <- c(0.420, 0.307, 0.496,  0.379, 0.511, 0.468 )

data.rs3087243.rep <- data.frame(region, A_freq, G_freq, AA_freq, AG_freq, GG_freq)
print(data.rs3087243.rep)
```

# rs3807306

The information about this genetic marker can be found here: 
https://www.ensembl.org/Homo_sapiens/Variation/Population?db=core;r=2:203873696-203874696;v=rs3087243;vdb=variation;vf=89583355

VCF: `7  128940626  rs3807306  G  A,C,T` 

```{r rs3807306_data, echo=FALSE}
region <- c("ALL", "AFR", "AMR", "EAS", "EUR", "SAS")
G_freq <- c(0.654 , 0.703 , 0.663 , 0.847 , 0.487, 0.553)
T_freq <- 1.00 - G_freq

GG_freq <- c(0.450, 0.498, 0.458,  0.726, 0.225, 0.325)
GT_freq <- c(0.409, 0.411, 0.409,  0.242, 0.525, 0.456)
TT_freq <- c(0.142, 0.091, 0.133,  0.032, 0.250, 0.219)

data.rs3807306.rep <- data.frame(region, G_freq, T_freq, GG_freq, GT_freq, TT_freq)
print(data.rs3807306.rep)
```

# rs1800629

The information about this genetic marker can be found here: 
https://www.ensembl.org/Homo_sapiens/Variation/Population?db=core;r=6:31574754-31575754;v=rs1800629;vdb=variation;vf=406243482

VCF: `6  31575254  rs1800629  G  A` 

```{r rs1800629_data, echo=FALSE}
region <- c("ALL", "AFR", "AMR", "EAS", "EUR", "SAS")
G_freq <- c(0.910  , 0.880  , 0.931  , 0.941  , 0.866 , 0.947 )
A_freq <- 1.00 - G_freq

GG_freq <- c(0.827, 0.773, 0.867,  0.885, 0.744, 0.898)
AA_freq <- c(0.008, 0.012, 0.006,  0.002, 0.012, 0.004)
AG_freq <- c(0.165, 0.215, 0.127,  0.113, 0.245, 0.098)

data.rs1800629.rep <- data.frame(region, G_freq, A_freq, GG_freq, AA_freq, AG_freq)
print(data.rs1800629.rep)
```

# rs1799724

The information about this genetic marker can be found here: 
https://www.ensembl.org/Homo_sapiens/Variation/Explore?db=core;r=6:31574205-31575205;v=rs1799724;vdb=variation;vf=406243312

VCF: `6  31574705  rs1799724  C  T` 

```{r rs1799724_data, echo=FALSE}
region <- c("ALL", "AFR", "AMR", "EAS", "EUR", "SAS")
C_freq <- c(0.901   , 0.976   , 0.817   , 0.875   , 0.906  , 0.881  )
T_freq <- 1.00 - C_freq

CC_freq <- c(0.814 , 0.953 , 0.657 ,  0.770 , 0.821 , 0.775 )
CT_freq <- c(0.174 , 0.045 , 0.320 ,  0.210 , 0.169 , 0.213 )
TT_freq <- c(0.012 , 0.002 , 0.023 ,  0.020 , 0.010 , 0.012 )

data.rs1799724.rep <- data.frame(region, C_freq, T_freq, CC_freq, CT_freq, TT_freq)
print(data.rs1799724.rep)
```

# Estimated distributions of genetic markers in healthy patients

In the beginning we have to prepare data. We are filtering data to get data about healthy patients and then remove missing data.
```{r}
data_HC <- data[data$Group == 'HC', ]
head(data_HC)

sum(is.na(data_HC)) 

data_HC <- na.omit(data_HC)
sum(is.na(data_HC))

dim(data_HC) 
```

Now we are going to create new variable for each allele and calculate distribution of them. Then we compare results with distributions from repository.

# rs2476601
```{r}
data.rs2476601 <- data_HC$rs2476601
unique(data.rs2476601)

table.rs2476601 <- table(data.rs2476601)
num.alleles.rs2476601 <- 2 * sum(table.rs2476601)
numA.alleles.rs2476601 <- sum(data.rs2476601=='AG')

proportion.allelesA.rs2476601 <- numA.alleles.rs2476601/num.alleles.rs2476601

proportion.rs2476601 <- data.frame(
  region = "sample",
  A_freq = proportion.allelesA.rs2476601,
  G_freq = 1 - proportion.allelesA.rs2476601
)

proportion.rs2476601
```
```{r}
data.plot <- rbind(data.rs2476601.rep[, c("region", "A_freq", "G_freq")], proportion.rs2476601)

data.plot <- pivot_longer(data.plot, A_freq:G_freq, values_to = "frequency", names_to = "allel")

rs2476601.plot <- ggplot(data=data.plot, aes(x=region, y=frequency, fill=allel)) +
  geom_bar(stat="identity", width=0.7)+
  scale_fill_discrete(labels=c("A", "G"))+
  ylab("data origin")+
  theme_minimal()
rs2476601.plot
```


# rs3087243

```{r}
data.rs3087243 <- data_HC$rs3087243
unique(data.rs3087243)

table.rs3087243 <- table(data.rs3087243)
num.alleles.rs3087243 <- 2 * sum(table.rs3087243)


numA.alleles.rs3087243 <- sum(data.rs3087243 == 'AA' | data.rs3087243 == 'AG')

proportion.allelesA.rs3087243 <- numA.alleles.rs3087243/num.alleles.rs3087243

proportion.rs3087243 <- data.frame(
  region = "sample",
  A_freq = proportion.allelesA.rs3087243,
  G_freq = 1 - proportion.allelesA.rs3087243
)

proportion.rs3087243
```

```{r}
data.plot <- rbind(data.rs3087243.rep[, c("region", "A_freq", "G_freq")], proportion.rs3087243)

data.plot <- pivot_longer(data.plot, A_freq:G_freq, values_to = "frequency", names_to = "allel")

rs3087243.plot <- ggplot(data=data.plot, aes(x=region, y=frequency, fill=allel)) +
  geom_bar(stat="identity", width=0.7)+
  scale_fill_discrete(labels=c("A", "G"))+
  ylab("data origin")+
  theme_minimal()
  
rs3087243.plot
```

# rs3807306

```{r}
data.rs3807306 <- data_HC$rs3807306
unique(data.rs3807306)

table.rs3807306 <- table(data.rs3807306)
num.alleles.rs3807306 <- 2 * sum(table.rs3807306)


numT.alleles.rs3807306 <- sum(data.rs3807306 == 'GT' | data.rs3807306 == 'TT')

proportion.allelesT.rs3807306 <- numT.alleles.rs3807306/num.alleles.rs3807306

proportion.rs3807306 <- data.frame(
  region = "sample",
  T_freq = proportion.allelesT.rs3807306,
  G_freq = 1 - proportion.allelesT.rs3807306
)

proportion.rs3807306
```
```{r}
data.plot <- rbind(data.rs3807306.rep[, c("region", "T_freq", "G_freq")], proportion.rs3807306)

data.plot <- pivot_longer(data.plot, T_freq:G_freq, values_to = "frequency", names_to = "allel")

rs3807306.plot <- ggplot(data=data.plot, aes(x=region, y=frequency, fill=allel)) +
  geom_bar(stat="identity", width=0.7)+
  scale_fill_discrete(labels=c("T", "G"))+
  ylab("data origin")+
  theme_minimal()
  
rs3807306.plot
```


# rs1800629
```{r}
data.rs1800629 <- data_HC$rs1800629
unique(data.rs1800629)

table.rs1800629 <- table(data.rs1800629)
num.alleles.rs1800629 <- 2 * sum(table.rs1800629)


numA.alleles.rs1800629 <- sum(data.rs1800629 == 'AG' | data.rs1800629 == 'AA')

proportion.allelesA.rs1800629 <- numA.alleles.rs1800629/num.alleles.rs1800629

proportion.rs1800629 <- data.frame(
  region = "sample",
  A_freq = proportion.allelesA.rs1800629,
  G_freq = 1 - proportion.allelesA.rs1800629
)

proportion.rs1800629
```
```{r}
data.plot <- rbind(data.rs1800629.rep[, c("region", "A_freq", "G_freq")], proportion.rs1800629)

data.plot <- pivot_longer(data.plot, A_freq:G_freq, values_to = "frequency", names_to = "allel")

rs1800629.plot <- ggplot(data=data.plot, aes(x=region, y=frequency, fill=allel)) +
  geom_bar(stat="identity", width=0.7)+
  scale_fill_discrete(labels=c("A", "G"))+
  ylab("data origin")+
  theme_minimal()
  
rs1800629.plot
```

# rs1799724
```{r}
data.rs1799724 <- data_HC$rs1799724
unique(data.rs1799724)

table.rs1799724 <- table(data.rs1799724)
num.alleles.rs1799724 <- 2 * sum(table.rs1799724)


numT.alleles.rs1799724 <- sum(data.rs1799724 == 'CT' | data.rs1799724 == 'TT')

proportion.allelesT.rs1799724 <- numT.alleles.rs1799724/num.alleles.rs1799724

proportion.rs1799724 <- data.frame(
  region = "sample",
  T_freq = proportion.allelesT.rs1799724,
  C_freq = 1 - proportion.allelesT.rs1799724
)

proportion.rs1799724
```

```{r}
data.plot <- rbind(data.rs1799724.rep[, c("region", "T_freq", "C_freq")], proportion.rs1799724)

data.plot <- pivot_longer(data.plot, T_freq:C_freq, values_to = "frequency", names_to = "allel")

rs1799724.plot <- ggplot(data=data.plot, aes(x=region, y=frequency, fill=allel)) +
  geom_bar(stat="identity", width=0.7)+
  scale_fill_discrete(labels=c("C", "T"))+
  ylab("data origin")+
  theme_minimal()
  
rs1799724.plot
```

# Genotype Distribution Between CFS and HC

To test if the genotype distribution between CFS and HC is significantly different we used the Chi-Squared test and the Fisher's exact test. According to https://www.ccjm.org/content/84/9_suppl_2/e20 if substantial (above 20%) of expected cell counts is less than 5, the Fisher's exact test should be used.

## rs2476601
```{r}
p.values <- c()

contingency_table <- table(data$Group, data$rs2476601)
contingency_table
fisher_test_result <- fisher.test(contingency_table)
fisher_test_result

p.values <- c(p.values,fisher_test_result$p.value)
```

# rs3087243
```{r}
contingency_table <- table(data$Group, data$rs3087243)
contingency_table
chi_square_result <- chisq.test(contingency_table)
chi_square_result

p.values <- c(p.values,chi_square_result$p.value)
```

# rs3807306
```{r}
contingency_table <- table(data$Group, data$rs3807306)
contingency_table
chi_square_result <- chisq.test(contingency_table)
chi_square_result

p.values <- c(p.values,chi_square_result$p.value)
```

# rs1800629
```{r}
contingency_table <- table(data$Group, data$rs1800629)
contingency_table
fisher_test_result <- fisher.test(contingency_table)
fisher_test_result

p.values <- c(p.values,fisher_test_result$p.value)
```

# rs1799724
```{r}
contingency_table <- table(data$Group, data$rs1799724)
contingency_table
chi_square_result <- chisq.test(contingency_table)
chi_square_result

p.values <- c(p.values,chi_square_result$p.value)
```

# Bonferroni's Correction
Now we can apply the Bonferroni's correction to adjust the p-values.
```{r}
adjusted_p_values <- p.adjust(p.values, method = "bonferroni")
adjusted_p_values


gene_names <- c("rs2476601", "rs3087243", "rs3807306", "rs1800629", "rs1799724")
tests_used <- c("Fisher's Exact Test", "Chi-squared Test", "Chi-squared Test", "Fisher's Exact Test", "Chi-squared Test")

result_table <- data.frame(
  Genetic_Marker = gene_names,
  Test_Used = tests_used,
  P_Value = p.values,
  Adjusted_P_Value = adjusted_p_values
)

result_table
```
As we can see, there is a significant difference in genotype distribution between CFS and HC for rs3087243 after Bonferroni correction. For all other genetic markers there are no significant differences in genotype distributions between the groups.

# Creating models

Next, we created models using "Group" as the outcome variable and adjusting for 'Gender'. 

## rs2476601
```{r}
new_data <- na.omit(data)
new_data$Group_binary <- as.factor(ifelse(new_data$Group == "CFS", 1, 0))

mod.rs2476601 <- glm(new_data$Group_binary~new_data$rs2476601+new_data$Gender, family=binomial(link='probit'))
summary(mod.rs2476601)
```

# rs3087243
```{r}
mod.rs3087243 <- glm(new_data$Group_binary~new_data$rs3087243+new_data$Gender, family=binomial(link='probit'))
summary(mod.rs3087243)
```

# rs3807306
```{r}
mod.rs3807306 <- glm(new_data$Group_binary~new_data$rs3807306+new_data$Gender, family=binomial(link='probit'))
summary(mod.rs3807306)
```

# rs1800629
```{r}
mod.rs1800629 <- glm(new_data$Group_binary~new_data$rs1800629+new_data$Gender, family=binomial(link='probit'))
summary(mod.rs1800629)
```

# rs1799724
```{r}
mod.rs1799724 <- glm(new_data$Group_binary~new_data$rs1799724+new_data$Gender, family=binomial(link='probit'))
summary(mod.rs1799724)
```

# Full Model
```{r}
mod.full <- glm(Group_binary~., family=binomial(link='probit'),data=subset(new_data, select = -Group))
summary(mod.full)
```

# Empty Model
```{r}
mod.null <- glm(Group_binary~1, family=binomial(link='probit'),data=subset(new_data, select = -Group))
summary(mod.null)
```

# Combined Model
```{r}
mod.new <- glm(new_data$Group_binary~new_data$rs3087243+new_data$rs1799724+new_data$Gender, family=binomial(link='probit'))
summary(mod.new)
```

# Accuracy
```{r}
predictions <- predict(mod.new, newdata = new_data, type = "response")
predicted_labels <- ifelse(predictions > 0.5, 1, 0)
accuracy <- mean(predicted_labels == new_data$Group_binary)
accuracy

roc_curve <- roc(new_data$Group_binary, predictions)
plot(roc_curve, main = "ROC Curve", col = "blue", lwd = 2)

auc_value <- auc(roc(new_data$Group_binary, predictions))
auc_value
```
Performance of the model is mediocre.  
